---
title: "10_reservoir_binding_properties"
author: "JR"
date: "5/29/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../util/intersect_functions.R")
source("../util/_setup.R")
library(GenomicRanges)
library(tidyverse)
library(effectsize)
```

# Loading in peak_occuence_df and peak_occurence_matrix

```{r}

peak_occurrence_matrix <- read.table("../01_consensus_peaks/results/lncrna_mrna_promoter_peak_occurence_matrix.tsv")

# Read in reservoir annotations combined with promoter peak occurence df. 
peak_occurrence_df <- read_csv("../08_defining_reservoirs/results/08_peak_occurrence_df_promoter_types.csv")

```


# We are setting up to look at the binding distribution on expressed lncRNA and mRNA vs reservoir mRNA lncRNA binding. 
#1 Distribution of reservoir vs all subsetted by lncRNA and mRNA
Figure 4C
```{r}
# Let's look at distribution of binding events to filter on those that have more than typical numbers of DBPs bound to promoter.
# first reverting back into a data frame with specific columns 
g <- ggplot(peak_occurrence_df, aes(x = number_of_dbp))
g + geom_histogram(binwidth = 5)  + 
  xlim(30, 100) +
  facet_wrap(expression ~ reservoir, scales = "free_y")

ggsave("figures/exp_not_exp_binding_histogram.png")
ggsave("figures/exp_not_exp_binding_histogram.pdf")

```
# We find that expressed or not expressed promoters still have similar distributions of DBP binding events. With some non-expressed promoters containing 100 independent DBP localization events. 


#2 Chi-Squared test for DBPs enriched and depleted at reservoirs versus non-reservoir promoters.
```{r}
# Subset columns to only the reservoirs
res_occurrence_matrix <- peak_occurrence_matrix[, peak_occurrence_df$gene_id[peak_occurrence_df$reservoir == T]]
# Non reservoirs that also have more than 7 peaks bound
non_res_occurrence_matrix <- peak_occurrence_matrix[,  peak_occurrence_df$gene_id[peak_occurrence_df$reservoir == F & peak_occurrence_df$number_of_dbp > 7]]


# Number of promoters bound in each category
dbp_binding <- data.frame(dbp = rownames(peak_occurrence_matrix),
                          "num_res_promoters_bound" = rowSums(res_occurrence_matrix),
                          "num_nonres_promoters_bound" = rowSums(non_res_occurrence_matrix))

dbp_binding <- dbp_binding %>%
  mutate("num_res_promoters_not_bound" = ncol(res_occurrence_matrix) - num_res_promoters_bound,
         "num_nonres_promoters_not_bound" = ncol(non_res_occurrence_matrix) - num_nonres_promoters_bound)



for(i in 1:nrow(dbp_binding)) {
  df1 <- data.frame("gene_type" = c("reservoir","reservoir", "non_res", "non_res"),
                    "promoter_bound" = c("bound", "not_bound", "bound", "not_bound"),
                    "count" = c(dbp_binding$num_res_promoters_bound[i],
                                dbp_binding$num_res_promoters_not_bound[i],
                                dbp_binding$num_nonres_promoters_bound[i],
                                dbp_binding$num_nonres_promoters_not_bound[i])) %>%
    pivot_wider(names_from = gene_type, values_from = count) %>%
    column_to_rownames("promoter_bound") %>%
    as.matrix()
  
  csres <- chisq.test(df1)
  phi_coef <- phi(df1)
  
  # Add the results to the data frame
  dbp_binding[i, "chisq_stat"] <- csres$statistic
  dbp_binding[i, "chisq_pval"] <- csres$p.value
  dbp_binding[i, "reservoir_peaks_expected"] <- csres$expected["bound", "reservoir"]
  dbp_binding[i, "phi_coefficient"] <- phi_coef$phi
}


#Adjusting P value with BH correction, observed - expected and writing this to a .csv file

dbp_binding$padj <- p.adjust(dbp_binding$chisq_pval, method = "BH")
dbp_binding$diff <- log2(dbp_binding$num_res_promoters_bound/dbp_binding$reservoir_peaks_expected)
write_csv(dbp_binding, "results/reservoir_chi_squared_results.csv")


#Plotting Chi-squared test results 
g <- ggplot(dbp_binding, aes(x = diff, y = -log10(chisq_pval), label = dbp))
g + geom_point() +
  geom_text() +
  geom_hline(yintercept = -log10(0.001))
ggsave("figures/chi-squared_enrichment_depletion.pdf")

# This shows that there are not any DBPs significantly enriched on reservoirs (P value < 0.01 and 2 fold enrichment)

```



#TODO ? UMAP reservoir promoter annotations ?






