---
title: "Reservoir binding properties"
author: "JR"
date: "5/29/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(tidyverse)
library(effectsize)
source("../util/intersect_functions.R")
source("../util/_setup.R")
```

# Loading in peak_occuence_df and peak_occurence_matrix

```{r}
base_path <- "../01_consensus_peaks/results"
peak_occurrence_matrix <- read.table(file.path(base_path,
                        "lncrna_mrna_promoter_peak_occurence_matrix.tsv"))

# Read in reservoir annotations combined with promoter peak occurence df. 
base_path <- "../08_defining_reservoirs/results"
peak_occurrence_df <- read.csv(file.path(base_path,
                                "08_peak_occurrence_df_promoter_types.csv"))
```

We are setting up to look at the binding distribution on expressed
lncRNA and mRNA vs reservoir mRNA lncRNA binding. 
Distribution of reservoir vs all subsetted by lncRNA and mRNA

```{r}
g <- ggplot(peak_occurrence_df, aes(x = number_of_dbp))
g + geom_histogram(binwidth = 5)  + 
  xlim(30, 100) +
  facet_wrap(expression~gene_type, scales = "free_y")
ggsave("figures/many_binders_histogram.png")
ggsave("figures/many_binders_histogram.pdf")

#### FIGURE: Figure 4C
g <- ggplot(peak_occurrence_df %>% filter(number_of_dbp > 7), 
            aes(x = number_of_dbp, 
                fill = factor(reservoir, labels = c("Non-res", "Reservoir"))))
g + geom_density(alpha = 0.2) + 
  scale_fill_manual(values = c("#424242","#a8404c"),
                    name = " ") + 
  facet_grid(~gene_type) +
  ggtitle("DBP binding density",
          subtitle = "Promoters w/ > 7 DBPs")
ggsave("figures/many_binders_density_plot.png")
ggsave("figures/many_binders_density_plot.pdf")


data_summary <- function(x) {
   m <- mean(x)
   ymin <- m - sd(x)
   ymax <- m + sd(x)
   return(c(y = m, ymin = ymin, ymax = ymax))
}

g <- ggplot(peak_occurrence_df %>% filter(number_of_dbp > 7), 
            aes(x = factor(reservoir, labels = c("Non-res", "Reservoir")),
                           y = number_of_dbp))
g + geom_violin() + stat_compare_means() +
  facet_grid(~gene_type) +
  stat_summary(fun.data = data_summary) +
  stat_summary(fun.data = function(x) {
    data.frame(y = 90, 
               label = paste("Mean=", round(mean(x),1))) },
               geom = "text")  +
  ggtitle("DBP binding density",
          subtitle = "Promoters w/ > 7 DBPs")

peak_occurrence_df %>% filter(number_of_dbp > 7) %>%
  group_by(reservoir, gene_type) %>%
  summarize(mean_dbps = mean(number_of_dbp)) %>%
  pivot_wider(names_from = reservoir,
              values_from = mean_dbps) %>%
  mutate(diff = `0` - `1`,
         fc = `1`/`0`)
```

We find that expressed or not expressed promoters still have similar 
distributions of DBP binding events. With some non-expressed promoters 
containing 100 independent DBP localization events. 


Chi-Squared test for DBPs enriched and depleted at reservoirs 
versus non-reservoir promoters.

```{r}
# Subset columns to only the reservoirs
res_gene_ids <- peak_occurrence_df$gene_id[peak_occurrence_df$reservoir == T]
res_occurrence_matrix <- peak_occurrence_matrix[ , res_gene_ids]
# Non reservoirs that also have more than 7 peaks bound
non_res <- peak_occurrence_df$gene_id[peak_occurrence_df$reservoir == F & 
                                        peak_occurrence_df$number_of_dbp > 7]
non_res_occurrence_matrix <- peak_occurrence_matrix[ , non_res]


# Number of promoters bound in each category
dbp_binding <- data.frame(dbp = rownames(peak_occurrence_matrix),
                          "num_res_promoters_bound" = rowSums(res_occurrence_matrix),
                          "num_nonres_promoters_bound" = rowSums(non_res_occurrence_matrix))

dbp_binding <- dbp_binding %>%
  mutate("num_res_promoters_not_bound" = ncol(res_occurrence_matrix) - num_res_promoters_bound,
         "num_nonres_promoters_not_bound" = ncol(non_res_occurrence_matrix) - num_nonres_promoters_bound)



for (i in 1:nrow(dbp_binding)) {
  df1 <- data.frame("gene_type" = c("reservoir","reservoir", "non_res", "non_res"),
                    "promoter_bound" = c("bound", "not_bound", "bound", "not_bound"),
                    "count" = c(dbp_binding$num_res_promoters_bound[i],
                                dbp_binding$num_res_promoters_not_bound[i],
                                dbp_binding$num_nonres_promoters_bound[i],
                                dbp_binding$num_nonres_promoters_not_bound[i])) %>%
    pivot_wider(names_from = gene_type, values_from = count) %>%
    column_to_rownames("promoter_bound") %>%
    as.matrix()
  
  csres <- chisq.test(df1)
  phi_coef <- phi(df1)
  
  # Add the results to the data frame
  dbp_binding[i, "chisq_stat"] <- csres$statistic
  dbp_binding[i, "chisq_pval"] <- csres$p.value
  dbp_binding[i, "reservoir_peaks_expected"] <- csres$expected["bound", "reservoir"]
  dbp_binding[i, "phi_coefficient"] <- phi_coef$phi
}


# Adjusting P value with BH correction, observed - expected and 
# writing this to a .csv file
dbp_binding$padj <- p.adjust(dbp_binding$chisq_pval, method = "BH")
dbp_binding$diff <- log2(dbp_binding$num_res_promoters_bound/dbp_binding$reservoir_peaks_expected)
write_csv(dbp_binding, "results/reservoir_chi_squared_results.csv")


# plotting Chi-squared test results 
g <- ggplot(dbp_binding, aes(x = diff, y = -log10(padj), label = dbp))
g + geom_point() +
  geom_hline(yintercept = -log10(0.001), lty = 2) +
  geom_vline(xintercept = -1, lty = 2) +
  geom_vline(xintercept = 1, lty = 2) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_text_repel(data = dbp_binding %>% filter(-log10(padj) > 10 &
                    diff < -1)) + 
  ggtitle("Reservoir DBP bias",
          subtitle = "Chi-squared test res vs. non-res")
ggsave("figures/reservoir_dbp_chisq.pdf")
ggsave("figures/reservoir_dbp_chisq.png")

g <- ggplot(dbp_binding, aes(x = diff, y = -log10(padj), label = dbp))
g + geom_point() +
  geom_hline(yintercept = -log10(0.001), lty = 2) +
  geom_vline(xintercept = -1, lty = 2) +
  geom_vline(xintercept = 1, lty = 2) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_text_repel(data = dbp_binding %>% filter(-log10(padj) > 3 & diff > 0)) + 
  ylim(0,5)
ggsave("figures/reservoir_dbp_chisq_enriched.pdf")
ggsave("figures/reservoir_dbp_chisq_enriched.png")
```

```{r}
res_dbp_sig <- dbp_binding %>% filter(padj < 0.001, diff < -1 | diff > 1) %>%
  group_by(diff < 0) %>%
  summarize(count = n())
knitr::kable(res_dbp_sig)
```


```{r reservoir-umap, message=FALSE}
# TODO ? UMAP reservoir promoter annotations ?
```








