---
title: "Expression vs. DNA-binding at promoters"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)

library(tidyverse)
library(GenomicRanges)
library(GenomicFeatures)
library(rtracklayer)
library(ggrepel)
library(ggpubr)
library(Rsubread)
library(regioneR)

source("../util/intersect_functions.R")
source("../util/_setup.R")
```


# Purpose: Compare expression of genes to binding in promoter regions
 
This script extracts expression counts from K562 RNA-seq data and identifies DNA-binding events at promoters from the K562 ChIP-seq consensus peak dataset. We have included further analysis on promoters that contain ≥7 DNA-binding events but are not expressed.


### Source for RNA-seq data

total RNA-seq for K562 dataset: https://www.encodeproject.org/experiments/ENCSR885DVH/


### Analysis of RNA-seq data
Loading in the RNAseq data and converting to TPM and saving as CSV of TPM values
```{r read_counts_to_TPM}
if(!file.exists("results/k562_tpm.csv")) {
  
  fl <- list.files("data/", pattern = ".bam", full.names = T)
  if(length(fl < 2)) {
    system("cd data;
           wget https://www.encodeproject.org/files/ENCFF625ZBS/@@download/ENCFF625ZBS.bam",
           intern = FALSE, ignore.stdout = TRUE, ignore.stderr = TRUE)
    system("cd data;
           wget https://www.encodeproject.org/files/ENCFF630HEX/@@download/ENCFF630HEX.bam",
           intern = FALSE, ignore.stdout = TRUE, ignore.stderr = TRUE)
  }
  
  counts <- featureCounts(c("data/ENCFF625ZBS.bam", "data/ENCFF630HEX.bam"),
                          annot.ext = "/Shares/rinn_class/data/genomes/human/gencode/v32/gencode.v32.annotation.gtf",
                          isGTFAnnotationFile = T,
                          isPairedEnd = TRUE,
                          nthreads = 16)
  
  rpk <- counts$counts / (counts$annotation$Length/1000)
  expression <- data.frame("rpk" = rpk) %>%
    rownames_to_column("gene_id") %>%
    pivot_longer(2:3, names_to = "sample", values_to = "rpk")
  
  expression_summary <- expression %>%
    group_by(sample) %>%
    summarize(total_rpk = sum(rpk, na.rm = T))
  expression_summary$rpk_m <- expression_summary$total_rpk / 1e6
  expression <- merge(expression, expression_summary)
  expression$tpm <-expression$rpk / expression$rpk_m
  tpm <- expression %>% group_by(gene_id) %>%
    summarize(tpm = mean(tpm, na.rm = T))
  write_csv(tpm, "results/k562_tpm.csv")
}
tpm <- read.csv("results/k562_tpm.csv")
```


# Let's add the expression data to our peak_occurence_df generated in 01_consensus_peaks.
# Our strategy will be to keep adding variables to our observations (promoters). 
# We can then plot from this data frame as it grows. 
```{r}
peak_occurrence_df <- read.csv("../01_consensus_peaks/results/peak_occurence_dataframe.csv")
expression_occurrence_df <- read.csv("results/peak_tpm_occurence_df.csv")
peak_occurrence_df <- merge(peak_occurrence_df, expression_occurrence_df)
#filtering out an old column used to be called number_tfs is now number_of_dbp.
peak_occurrence_df <- peak_occurrence_df %>% dplyr::select(-number_of_tfs)

#Adding expression to the new peak occurence data frame nomenclature that will be used as a starting point in 08_defining_reservoirs.

write_csv(peak_occurrence_df, "results/peak_occurrence_df_exp_added.csv")
```

#### Plotting binding of lncRNAs and mRNAs versus expression

We see an upward trend in both lncRNA and mRNA promoters between number of binding events and expression output

```{r, message=FALSE, warning=FALSE}
### FIGURE: Figure 4A
## NOTE: The GAM smoothing (s) uses a shrinkage version of the cubic regression spline.
## The dip in lncRNA may be a result of the smoothing basis, but I still need to figure
## out how this is determined theoretically and whether it makes sense in this context. 
g <- ggplot(peak_occurrence_df, aes(y = log2(tpm+0.001), x = number_of_dbp, color = gene_type))
g + geom_point(data = peak_occurrence_df %>% filter(tpm < 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM)))
ggsave("figures/k562_expression_vs_lncRNA_mRNA_promoter_binding.png")
ggsave("figures/k562_expression_vs_lncRNA_mRNA_promoter_binding.pdf")
```

# Supplemental Figure 4a all promoters together versus expression

```{r, message=FALSE, warning=FALSE}
g <- ggplot(peak_occurrence_df, aes(y = log2(tpm+0.001), x = number_of_dbp))
g + geom_point(data = peak_occurrence_df %>% filter(tpm < 0.001), 
               shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"),
              color = "#424242") +
  scale_x_continuous(expand = c(0,0)) + 
  ggtitle("Expression vs. promoter binding events",
          subtitle = "lncRNA and mRNA promoters") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM)))
ggsave("figures/k562_expression_vs_all_promoter_binding.png")
ggsave("figures/k562_expression_vs_all_promoter_binding.pdf")
```

FIGURE 4B
Here we will plot the binding events as a funciton of off, low, medium and high expression
We find that at off and low expressed promoters there is no difference in bding events (including hundreds at a single promoter). 

```{r, message=FALSE, warning=FALSE}
## let's try a binned approach
quantile(peak_occurrence_df$tpm)
peak_occurrence_df$expression <- cut(peak_occurrence_df$tpm, breaks = c(-1, 0.001, 0.137, 3, 5000),
                                    labels = c("off", "low", "med", "high"))
table(peak_occurrence_df$expression)

g <- ggplot(peak_occurrence_df, aes(x = expression, y = number_of_dbp, color = gene_type))
g + geom_boxplot() + stat_compare_means() + 
  scale_color_manual(values = c("#a8404c", "#424242"))  
ggsave("figures/bin-exp_binding_vs_expression_density.png")
ggsave("figures/bin-exp_binding_vs_expression_density.pdf")
```


Figure 4C : Browser examples 


















############################################
Other ways of plotting the same stuff above.
############################################



Plotting Number of DBPs versus expression on mRNA promoters as a "heat map" Note the distribution around 30-60 DBPs bound at a given mRNA promoter
```{r}
# mRNA density plot
g <- ggplot(peak_occurrence_df %>% filter(gene_type == "protein_coding"), aes(y = log10(tpm), x = number_of_tfs, color = gene_type))
g + stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE, n = 500) + 
  scale_fill_gradientn(colors = colorRampPalette(colors = c("#ffffff", "#a8404c"))(100)) + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_continuous(expand = c(0,0)) + 
  ggtitle("mRNA binding vs. expression") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[10](TPM)))
ggsave("figures/mRNA_binding_vs_expression_density.png")
ggsave("figures/mRNA_binding_vs_expression_density.pdf")
```


Plotting Number of DBPs versus expression on lncRNA promoters as a "heat map" Note the faint distribution around 30-60 DBPs bound at a given lncRNA promoter
```{r}
# lncRNA density plot
g <- ggplot(peak_occurrence_df %>% filter(gene_type == "lncRNA"), aes(y = log10(tpm), x = number_of_tfs, color = gene_type))
g + stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE, n = 500) + 
  scale_fill_gradientn(colors = colorRampPalette(colors = c("#ffffff", "#a8404c"))(100)) + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_continuous(expand = c(0,0)) + 
  ggtitle("lncRNA binding vs. expression") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[10](TPM)))
ggsave("figures/lncRNA_binding_vs_expression_density.png")
ggsave("figures/lncRNA_binding_vs_expression_density.pdf")
```

Plotting lncRNA and mRNA density plots together from above, this is consisten with lncRNA promoters being depleted in general of DBPs similar to what was observed in Figure 2C
```{r}
# Plotting lncRNA and mRNA density together.
g <- ggplot(peak_occurrence_df, aes(y = log2(tpm), x = number_of_tfs))
g + geom_hex(bins =60) + facet_grid(~gene_type) + 
  scale_fill_gradientn(colors = colorRampPalette(colors = c("#ffffff", "#a8404c"))(100)) + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_continuous(expand = c(0,0))
ggsave("figures/lncRNA-mRNA_binding_vs_expression_density.png")
ggsave("figures/lncRNA-mRNA_binding_vs_expression_density.pdf")
```
























#############





### Chi-square statistics for reservoir promoters with POLII binding vs. without POLII binding

```{r chi_squared, message=FALSE, warning=FALSE}

# Get reservoir promoters
reservoir_promoters <- promoters[promoters$gene_id %in% reservoirs$gene_id,]

# Get peaks
poliia <- rtracklayer::import("/Shares/rinn_class/data/CLASS_2020/analysis/01_consensus_peaks/results/consensus_peaks/POLR2A_consensus_peaks.bed")
poliib <- rtracklayer::import("/Shares/rinn_class/data/CLASS_2020/analysis/01_consensus_peaks/results/consensus_peaks/POLR2B_consensus_peaks.bed")
polii_peaks <- c(poliia, poliib)

# Get reservoirs with and without POLII in them
pol_reservoir_overlaps <- findOverlaps(reservoir_promoters, polii_peaks)
pol_reservoir_proms <- reservoir_promoters[unique(pol_reservoir_overlaps@from)]
nn <- c(1:length(reservoir_promoters))
mm <- subset(nn, !(nn %in% pol_reservoir_overlaps@from))
nopol_reservoir_proms <- reservoir_promoters[mm]

pol_matrix <- count_peaks_per_feature(pol_reservoir_proms, peak_list, type = "occurrence")
nopol_matrix <- count_peaks_per_feature(nopol_reservoir_proms, peak_list, type = "occurrence")

pol_num_overlaps <- rowSums(pol_matrix)
nopol_num_overlaps <- rowSums(nopol_matrix)
num_peaks <- sapply(peak_list, length)
total_pol <- length(pol_reservoir_proms)
total_nopol <- length(nopol_reservoir_proms)
total_reservoirs <- length(reservoir_promoters)

chisq_res <- data.frame("tf" =character(),
                  "polii_peaks_observed" = numeric(),
                  "chisq_stat" = numeric(),
                  "chisq_pval" = numeric(),
                  "polii_peaks_expected" = numeric())

for(i in 1:length(num_peaks)) {
  total_peaks <- num_peaks[[i]]
  df1 <- data.frame("gene_type" = c("pol","pol", "nopol", "nopol"),
                    "promoter_bound" = c("bound", "not_bound", "bound", "not_bound"),
                    "count" = c(pol_num_overlaps[[i]],
                                total_pol - pol_num_overlaps[[i]],
                                nopol_num_overlaps[[i]],
                                total_nopol - nopol_num_overlaps[[i]])) %>%
    pivot_wider(names_from = gene_type, values_from = count) %>%
    column_to_rownames("promoter_bound") %>%
    as.matrix()
  csres <- chisq.test(df1)
  tdf <- data.frame("tf" = names(num_peaks[i]),
                    "polii_peaks_observed" = pol_num_overlaps[[i]],
                    "chisq_stat" = csres$statistic,
                    "chisq_pval" = csres$p.value,
                    "polii_peaks_expected" = csres$expected[1,1])
  chisq_res <- bind_rows(chisq_res, tdf)
}
chisq_res$padj <- p.adjust(chisq_res$chisq_pval, method = "BH")
chisq_res$diff <- chisq_res$polii_peaks_observed - chisq_res$polii_peaks_expected

write_csv(chisq_res, "results/k562_polII_reservoirs_chisquare.csv")
```


Figure C: Histogram of promoters by number of DNA-binding factors bound
```{r binding_histogram, message=FALSE, warning=FALSE}

peak_occurrence_df$expression <- "off"
peak_occurrence_df[which(peak_occurrence_df$tpm > 0.001), "expression"] <- "expressed"

g <- ggplot(peak_occurrence_df, aes(x = number_of_tfs))
g + geom_density(alpha = 0.2, fill = "#424242")  +
  geom_vline(xintercept = 30, lty = 2) +
  geom_vline(xintercept = 100, lty = 2)+
  ggtitle("Promoter Density vs Number of TFs")

ggsave("figures/k562_binding_histogram.png")
ggsave("figures/k562_binding_histogram.pdf")

```


Figure D: Histogram of promoters by number of DNA-binding factors bound for different biotypes and expression levels
```{r expression_histogram, message=FALSE, warning=FALSE}

g <- ggplot(peak_occurrence_df, aes(x = number_of_tfs))
g + geom_histogram(binwidth = 5)  + 
  xlim(30, 100) +
  facet_wrap(expression~gene_type, scales = "free_y")
  
ggsave("figures/k562_binding_histogram_expression.png")
ggsave("figures/k562_binding_histogram_expression.pdf")

```


Figure E: Histogram of promoters by gene expression for different DNA-binding factor densities. In red are prommoters with fewer than 25 factors bound). In black are promomters with 25 or more factors bound.
```{r plot_peak_occurrence, message=FALSE, warning=FALSE}
peak_occurrence_df$tf_binding <- "low"
peak_occurrence_df[which(peak_occurrence_df$number_of_tfs > 25), "tf_binding"] <- "high"

g <- ggplot(peak_occurrence_df, aes(x = log10(tpm), color = tf_binding))
g + geom_density(alpha = 0.2) +
  scale_color_manual(values = c("#424242","#a8404c"), name = "tf_binding")
ggsave("figures/k562_expression_histogram.png")
ggsave("figures/k562_expression_histogram.pdf")
```


Figure F: Chi-square values vs. difference in expected and observed binding for DNA-binding factors at promoters with or without POLII bound
```{r plot_chi_square, message=FALSE, warning=FALSE}
g <- ggplot(chisq_res, aes(x = diff, y = -log10(padj)))
g + geom_point(colour = "#424242") +
    geom_text_repel(
    data = subset(chisq_res, diff > 100 | diff < -10),
    aes(label = tf),
    size = 3,
    color ="#a8404c",
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")) +
  geom_vline(xintercept = 0, lty = 2) +
  ggtitle("Reservoirs: polII vs non-polII") 

ggsave("figures/k562_polII_reservoirs_chisquare.png")
ggsave("figures/k562_polII_reservoirs_chisquare.pdf")
```


## THIS IS FOR umap_with_metadata.
# Median expression of bound promoters per TF

```{r}
binding_df <- promoter_peak_occurence %>%
  as.data.frame() %>%
  rownames_to_column("tf") %>%
  pivot_longer(cols = 2:ncol(.), names_to = "gene_id", values_to = "bound") %>%
  filter(bound > 0)
binding_df <- merge(binding_df, rnaseq_exp, all.x = T)


binding_summary <- binding_df %>% group_by(tf) %>%
  summarize(mean_expression = mean(tpm, na.rm = T),
            median_expression = median(tpm, na.rm = T),
            sd_expression = sd(tpm, na.rm = T),
            cv_expression = (sd_expression/mean_expression)*100,
            num_promoters_bound = n(),
            promoters_bound = paste(gene_id, collapse = ";"))

binding_summary <- binding_summary %>% arrange(-cv_expression)
binding_summary$index <- 1:nrow(binding_summary)

g <- ggplot(binding_summary, aes(x = index, y = cv_expression, label = tf))
g + geom_point() + geom_text()

write_csv(binding_summary, "results/expression_summary_per_tf.csv")
```

```


